---
title: "R Notebook"
output: github_document
---

## Definition of Forecasting

Forecasting in time series analysis refers to the process of making predictions about future values based on previously observed values. It involves using historical patterns, trends, and relationships within the data to project future outcomes.

### Deterministic Approaches

Deterministic forecasting models assume that future values are completely determined by past values, with no random components. These approaches:

- Rely on mathematical equations that precisely define the relationship between variables
- Do not account for randomness or uncertainty in predictions
- Work well for systems with clear, stable patterns and minimal noise
- Include methods like linear regression, exponential smoothing, and decomposition techniques


### 1. Linear Trend Model 
```{r}
# Generate sample time series with linear trend
set.seed(123)
t <- 1:100
y <- 10 + 0.5*t + rnorm(100, 0, 5)
ts_data <- ts(y)

# Fit linear trend model
time_index <- time(ts_data)  # Extract time index
linear_model <- lm(ts_data ~ time_index)  # Use correct time variable

# Forecast next 10 periods
new_time <- data.frame(time_index = seq(101, 110))  # Ensure column name matches

# Predict future values
forecast_values <- predict(linear_model, newdata = new_time)

# Plot results
plot(ts_data, xlim = c(1, 110), ylim = c(min(y, forecast_values), max(y, forecast_values)),
     main = "Linear Trend Forecast")
lines(101:110, forecast_values, col = "red", lwd = 2)

```


### 2. Polynomial Trend Model

```{r}

# Generate sample time series with quadratic trend
set.seed(123)
t <- 1:100
y <- 10 + 0.5*t + 0.02*t^2 + rnorm(100, 0, 5)
ts_data <- ts(y)

# Fit quadratic trend model
time_index <- time(ts_data)  # Extract time index explicitly
quad_model <- lm(ts_data ~ time_index + I(time_index^2))  # Quadratic regression

# Forecast next 10 periods
new_time <- data.frame(time_index = 101:110)  # Ensure column name matches model
forecast_values <- predict(quad_model, newdata = new_time)

# Plot results
plot(ts_data, xlim = c(1, 110), ylim = c(min(y, forecast_values), max(y, forecast_values)),
     main = "Quadratic Trend Forecast")
lines(101:110, forecast_values, col = "red", lwd = 2)

```


### 3. Simple Exponential Smoothing (SES) in Time Series Forecasting


```{r}
library(forecast)

# Generate sample time series without trend or seasonality
set.seed(123)
y <- 100 + rnorm(100, 0, 10)
ts_data <- ts(y)

# Apply simple exponential smoothing
ses_model <- ses(ts_data, h = 10)  # h is the forecast horizon
summary(ses_model)

# Plot results
plot(ses_model, main = "Simple Exponential Smoothing Forecast")
```



### 4. Holt’s Linear Method (Double Exponential Smoothing) in Time Series Forecasting

```{r}
# Generate sample time series with trend
set.seed(123)
t <- 1:100
y <- 100 + 0.5*t + rnorm(100, 0, 5)
ts_data <- ts(y)

# Apply Holt's method
holt_model <- holt(ts_data, h = 10)
summary(holt_model)

# Plot results
plot(holt_model, main = "Holt's Linear Method Forecast")
```



### 5. Holt-Winters Method (Triple Exponential Smoothing) in Time Series Forecasting

```{r}
# Generate sample time series with trend and seasonality
set.seed(123)
t <- 1:100
trend <- 0.2*t
seasonal <- 15*sin(2*pi*t/12)  # Period of 12
y <- 100 + trend + seasonal + rnorm(100, 0, 5)
ts_data <- ts(y, frequency = 12)

# Apply Holt-Winters method
hw_model <- hw(ts_data, seasonal = "additive", h = 24)
summary(hw_model)

# Plot results
plot(hw_model, main = "Holt-Winters Method Forecast")
```



### 6. Classical Decomposition in Time Series Analysis

```{r}
# Generate sample time series
set.seed(123)
t <- 1:120
trend <- 0.3*t
seasonal <- 20*sin(2*pi*t/12)
y <- trend + seasonal + rnorm(120, 0, 5)
ts_data <- ts(y, frequency = 12)

# Classical decomposition
decomp <- decompose(ts_data)
plot(decomp)

# Extract time index and remove NA values from trend component
time_index <- time(ts_data)
valid_trend <- na.omit(decomp$trend)
valid_time <- time_index[!is.na(decomp$trend)]  # Matching time values

# Fit trend model
trend_model <- lm(valid_trend ~ valid_time)

# Forecast next 12 months
future_time <- data.frame(valid_time = seq(121, 132))
trend_forecast <- predict(trend_model, newdata = future_time)

# Handling NA values in trend forecast
if (any(is.na(trend_forecast))) {
  trend_forecast[is.na(trend_forecast)] <- tail(valid_trend, 1)
}

# Extract seasonal pattern (repeating every 12 months)
season_forecast <- decomp$seasonal[1:12]

# Compute final forecast
forecast_values <- trend_forecast + season_forecast

# Plot results
plot(ts_data, xlim = c(1, 132), ylim = range(c(y, forecast_values), na.rm = TRUE),
     main = "Classical Decomposition Forecast")
lines(121:132, forecast_values, col = "red", lwd = 2)

```



### 7. STL (Seasonal and Trend decomposition using LOESS) in Time Series Analysis
```{r}
library(forecast)

# Generate sample time series
set.seed(123)
t <- 1:120
trend <- 0.3*t
seasonal <- 20*sin(2*pi*t/12)
y <- trend + seasonal + rnorm(120, 0, 5)
ts_data <- ts(y, frequency = 12)

# STL decomposition
stl_decomp <- stl(ts_data, s.window = "periodic")
plot(stl_decomp)

# Forecast using STL
stl_forecast <- forecast(stl_decomp, method = "naive", h = 24)
plot(stl_forecast, main = "STL Decomposition Forecast")
```



### 8. Singular Spectrum Analysis (SSA) in Time Series Analysis

```{r}
library(Rssa)

# Generate sample time series
set.seed(123)
t <- 1:100
trend <- 0.2*t
seasonal <- 15*sin(2*pi*t/12)
y <- trend + seasonal + rnorm(100, 0, 3)
ts_data <- ts(y)

# Apply SSA
ssa_decomp <- ssa(ts_data, L = 48)
plot(ssa_decomp)

# Reconstruct components
groups <- list(trend = 1, seasonal = c(2,3), noise = c(4:10))
recon <- reconstruct(ssa_decomp, groups = groups)
plot(recon)

# Forecast
ssa_forecast <- forecast(ssa_decomp, groups = list(1:3), h = 24, 
                        method = "recurrent")
plot(ssa_forecast)
```


### 9. Centered Moving Average (CMA)

```{r}
# Generate sample time series
set.seed(123)
y <- rnorm(100, 0, 1)
for(i in 2:100) {
  y[i] <- 0.7*y[i-1] + rnorm(1, 0, 1)
}
y <- y + 10 + 0.1*(1:100)
ts_data <- ts(y)

# Apply moving average
ma <- stats::filter(ts_data, filter = rep(1/12, 12), sides = 2)

# Plot results
plot(ts_data, main = "Moving Average Smoothing")
lines(ma, col = "red")

# Use for forecasting
forecast_values <- c(ma[length(ma)-11:0])
for(i in 1:12) {
  next_val <- mean(c(tail(ts_data, 11), forecast_values[1:i]))
  forecast_values <- c(forecast_values, next_val)
}

lines(101:112, forecast_values[13:24], col = "blue")
```
